Evaluating machine translation for assimilation

### 2015-02-18 2015-eamt-assim/2015-eamt-assim.tex
Added the abstract, intro, methodology, setup and bibliography
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58988 72bbbca6-d526-0410-a7d9-f06f51895060

\usepackage{multirow}

@@

\title{A gap-filling approach to evaluation of machine translation for assimilation}

\author{First Author\\
  Affiliation / Address line 1\\
  Affiliation / Address line 2\\
  Affiliation / Address line 3\\
  {\tt email@domain}  \And
  Second Author\\
  Affiliation / Address line 1\\
  Affiliation / Address line 2\\
  Affiliation / Address line 3\\
  {\tt email@domain}}

@@

\begin{document}
\maketitle
\begin{abstract}
   ... 
\end{abstract}

@@

As suggested by \newcite{church93}, modern machine translation (MT) systems may
be divided into two broad categories according to their purpose: post-editing systems, e.g.
Google Translate 1 , and assimilation systems, e.g. Apertium 2 . Both kinds may be
systematically evaluated, either to control for quality in the development process or to
compare the systems. The evaluation methods, however, differ depending on the type of
the system. The methods are described in detail in \newcite{church93}.
In this work I perform the evaluation of an assimilation MT system called Apertium.
Apertium is an open-source community-maintained project that provides translation for
39 language pairs. The goal of this study is to assess whether machine translation
performed by this system assists users in understanding the gist of texts in an unknown
language. The evaluation is performed in three separate experiments for the following
translation directions: Basque to Spanish, English to Kazakh, and Tatar to Russian.
I build on the methodology established in \newcite{trosterud12} and
\newcite{oregan2013}, which are the works describing previous evaluations of
Apertium language pairs. \newcite{oregan2013} proposes the gap-filling task as a
means of assessing participant’s understanding of text. The same methodology is used in
this work and described in detail in the section below.

\section{Experiment design}

\subsection{Methodology}

This experiment is based on an idea that understanding of text may be assessed through
gap-filling questions. If the participant is able to restore principal information that had
been left out from the text, they are assumed to have understood it.
During the course of the experiment, the participants are offered sentences in their native
language (the target language of the pair), with some key words left out. Each sentence
has a counterpart in the source language of the pair. I use four different task settings:
\begin{itemize}
\item  Target sentence only. The participants are asked to fill the gaps not being given
any context. This task serves as a baseline score and as an indicator of phrases
that can be completed using common knowledge or language intuition (e.g.
idioms and strong collocations);
\item  Target sentence and source sentence. In addition to the question, the participants
are offered the same sentence in the source language of the pair. By setup, the
participants have no command of this language, however, it may help them to fill
in proper nouns or loan words;
\item  Target sentence and machine-translated hint. In addition to the task, the
participants see the source sentence translated via Apertium. This type of task is
used for measuring the contribution of machine translation to understanding the
gist of text;
\item  Target sentence and both the source and machine-translated hint. This task is
added mostly for completing the possible combination of task types, and to check
for any unexpected insights, should they arise.
\end{itemize}

\subsection{Material}

The data are collected separately for each experiment. A set of sentences is drawn from
parallel corpora. The source sentences are translated into target language via Apertium.
The target sentences are processed with a special script that removes significant words.
For the current experiments, the word is removed if it falls into a list of allowed parts of
speech:
\begin{itemize}
\item nouns;
\item proper nouns;
\item adjectives;
\item adverbs;
\item lexical verbs.
\end{itemize}
Parts of speech are determined automatically with Apertium morphological analyser. To
rule out homonymy, the set of all possible POS attributions of a given word is checked
against the POS list. If all of the possible word analyses match the parts of speech listed
above, it is considered a candidate for removal.
Each sentence in the set is evaluated in three “gap densities” (ratio of gaps to words in the
sentence): 10\%, 20\% and 30\%. After the script finds all possible keyword candidates in
the sentence, it removes the specified percentage of words, evenly distributing the gaps
across the sentence.
To exclude headlines and also to allow at least one gap in 10\%-gap sentences, only the
sentences over 10 words long are used in the tasks.
The sentences for each experiment are drawn from various parallel text sources:
\begin{enumerate}
\item  For Basque-Spanish, from the corpus of legal texts Memorias de traducción del
Servicio Oficial de Traductores del IVAP 3
\item  For English-Kazakh, from the official website of the President of the Republic of
Kazakhstan 4
\item  For Tatar-Russian, from the following sources on three different topics:
  \begin{enumerate}
    \item  Casual conversations, from a textbook 5 of spoken Tatar;
    \item  Legal texts, from the Constitution and laws 6 of Tatarstan;
    \item  News, from the President of Tatarstan website 7 .
  \end{enumerate}
\end{enumerate}

Each set features 36 pairs of sentences. For the first two experiments the pairs are drawn
randomly from the corpora, for Tatar-Russian, compiled by hand by the developer of the
language pair. After processing, the material is transformed into XML used for uploading
into the evaluation system (see section 2.4).

\subsection{Participants}

For each experiment we called for native speakers of L2 of the language pair (i.e.
Spanish, Kazakh, and Russian) who had no command of L1 of the pair (Basque, English
and Tatar, respectively). We had 11 participants in Basque-Spanish evaluation, 8 in
English-Kazakh, and 28 in Tatar-Russian (although not everyone has completed their task
in full, see discussion).

@@

The evaluations take place online, in a system 8 designed specifically for various MT
evaluation tasks. We adapted the code to accommodate for the gap-filling tasks. The
tasks are uploaded into the system and manually distributed between the participants by
the following rules:
\begin{enumerate}
\item  Each participant evaluates every sentence (understood as a succession of words),
a total of 36;
\item  Each participant evaluates 9 sentences in each of the four modes (section 2.1);
\item  All sentences of the set are evaluated with 10, 20 and 30\% of words removed;
\item  Each sentence-mode-percentage combination is evaluated by at least three
\end{enumerate}
The participants are given the instructions in their native language above each task. The
tasks are split into smaller sets for the participants’ convenience. The instructions are to

read all the available hints and fill each gap with one suitable word, guessing if unsure.
Participants’ answers are recorded and marked correct or incorrect automatically. In
addition, the time taken to fill the gaps in one sentence is recorded.
Since gap-filling task requires open answers, it is possible that the participants may
provide words that fit the gaps well, but do not match the original answer. To account for
these cases, we process all the answers to detect possible synonyms. An answer is
considered a candidate synonym if given by two or more evaluators and does not match
the answer key. We record each candidate synonym along with the answer key and the
context sentence. Based on this data, a native speaker of the target language decides
whether the candidate synonym is indeed synonymic to the answer key in the given
context. We then check participants’ results against the compiled synonym list and
increase scores where appropriate.

@@

The following trends may be derived from the data above:
\begin{enumerate}
\item  In general, tasks with MT assistance score higher than task without MT
assistance;
\item  The more words are removed from context, the more difficult it is to fill the gaps
(see below for exceptions);
\item  Task without hints take less time to complete than others.
\end{enumerate}
The first two trends align with our expectations. Higher scores on MT tasks indicate
better understanding of text. The large number of gaps in the sentence makes it more
difficult to predict the answer based on the context, and also leaves more room for
translation mistakes. We would expect that tasks without hints take the largest amount of
time rather than that the smallest, since the participant must come up with their own
answer unassisted. However, in the no-hint task the participant reads only one sentence,
as opposed to two or three in other tasks, which may be the source of difference.
In Basque-Spanish and Tatar-Russian these trends are defined reasonably well. In
English-Kazakh, however, the scores for source-hint task (a parallel sentence in English
for assistance) are systematically higher than for other types of tasks. In addition, a 20\%no-hint task stands out with 60\% accuracy, which is substantially higher than any other
scores for this experiment.
Having analyzed the tasks and the responses for English-Kazakh, I have found the
following possible reasons of the obtained results:
\begin{enumerate}
\item  The experiment was planned for 11 evaluators, although only 8 attempted the
tasks and 6 completed them in full. Because of this, the sentences were evaluated
once or twice, and some were not attempted at all. To reliably compare results
across different categories, all tasks must be completed;
\item  Due to an error during experiment design, some sentences are not repeated across
categories, which also suggests that we may not compare these percentages;
\item  The outlier score of 60\% is made up by three sentences. Two of these scored
unusually high. It appeared that one of them is a news headline, and the other
contains clichés such as “express condolences”. There may also be interference
from other tasks of the set, because some of the sentences are drawn from the
same news article. We would expect the same sentences to contribute to higher
percentage in other task modes, but they were not completed there;
\item  It may be the case that the participants have at least some command of English,
which would have helped them to derive extra information in the source-hint
\end{enumerate}
It remains questionable whether we can compare results for different gap densities. In the
Basque-Spanish and Tatar-Russian experiments the 10, 20 and 30\% sets were comprised
of the same sentences. However, in each case different words were removed. It appears
that some content words are easier to fill than the others. This may explain why in
Basque-Spanish the 20\% MT tasks are completed with better accuracy than 10\% tasks.
It may be worth noting that the majority of the participants reported feeling frustrated in
the course of their work, especially while working on the no-hint tasks. 9 out of 51
participants quit the experiment before completing it.

>>>>

\usepackage{multirow}
\usepackage{natbib}

@@

\title{Evaluating machine translation for assimilation}

\author{Ekaterina Ageeva\\
  School of Linguistics\\
  Higher School of Economics\\
  Moscow, Russia\\
  {\tt evageeva\_2@edu.hse.ru}\\[2ex]
  \textbf{Francis M. Tyers}\\
  HSL-fakultetet\\ 
  UiT Norgga\'{a}rktala\v{s} universitehta \\
  9017 Romsa, Norway \\
  {\tt francis.tyers@uit.no}
  \And
  Mikel L. Forcada\\
  Dept. Llenguatges i Sistemes Inform\`{a}tics\\
  Universitat d'Alacant, Spain \\
  {\tt mlf@dlsi.ua.es}  \\[2ex]
  \textbf{Juan Antonio P\'{e}rez-Ortiz} \\
  Dept. Llenguatges i Sistemes Inform\`{a}tics\\
  Universitat d'Alacant, Spain \\
  {\tt japerez@dlsi.ua.es}
}

@@

\begin{document}
\maketitle
%Katya: has not it first been used in trosterud12?
\begin{abstract}
This paper provides additional observations on the viability of a strategy proposed in 2012 for evaluation of machine translation for assimilation purposes. The evaluation method is applied to three translation directions. To reduce preparation time, an automatic open-source task management system is introduced.
\end{abstract}

@@

As suggested by \citet{church93}, modern machine translation (MT) systems may
be divided into two broad categories according to their purpose: post-editing systems, e.g.
Google Translate\footnote{translate.google.com}, and assimilation systems, e.g. Apertium\footnote{apertium.org}. Both kinds may be systematically evaluated, either to control for quality in the development process or to compare the systems. Importantly, according to \citet{church93}, the evaluation methods must closely consider the system's primary purpose.

%\Katya: I'd like a citation for the "most massive application", could you suggest something?
Despite the fact that assimilation (or gisting) is currently the most massive application of MT, few methodologies are established for assimilation evaluation of MT. The methods include post-editing and comparison by bilingual experts \citep{ginesti09}, and multiple choice tests \citep{jones07}, \citep{trosterud12} These approaches are often costly and prone to subjectivity, see the discussion in \citep{oregan13}. As an alternative, the modification of Cloze test was introduced for assimilation evaluation, first in \citep{trosterud12} as a supplementary, and then in \citep{oregan13} as a stand-alone method. Prior to this, Cloze tests have been used to evaluate raw MT quality, see e.g. \citep{vanslype79}, \citep{somers00}. As a principal difference, \citet{trosterud12} and \citet{oregan13} propose to fill the gaps in the reference (human) translation.

The gap-filling method has been successfully used to evaluate the Basque-English Apertium language pair. In this work we extend the evaluation to three more language pairs: Basque-Spanish, English-Kazakh and Tatar-Russian. For the latter two pairs, the evaluation served as a quality check in the period of active development during the Google Summer of Code'14 program. In addition to evaluating, we explore new aspects of the experiment: the correlation between evaluators' scores, and the influence of the texts' linguistic domain. To facilitate the evaluation, we introduce an automated task creation system. This system is integrated into the Appraise MT evaluation platform \citep{federmann12}; the code is open source and is available on Github\footnote{https://github.com/Sereni/Appraise}.

We anticipate that MT systems under investigation will contribute to the users' understanding of text. We also expect to see different results depending on text domain and on the number of words left out from the task.

The paper is organized as follows: in section~\ref{sec:methodology} we describe the gap-filling method for assimilation evaluation: the task layout, the choice of words, and how the tasks are generated. Section~\ref{sec:setup} introduces the experimental material, the evaluators, the distribution of tasks and the evaluation procedure. In section~\ref{sec:results} we describe and discuss the experiment results. Finally, section~\ref{sec:conclusion} draws conclusions.

\section{Methodology}
\label{sec:methodology}

This section discusses the reasoning behind the gap-filling method and task structure. The gap-filling method of evaluating machine translation for assimilation purposes is based on the following notion: readers' understanding of a text correlates with the number of keywords they are able to correctly restore in this text. Therefore, the base of an assimilation task is a (reference) sentence, where some of the keywords are blacked out, or removed. The sentence is produced by a human (as opposed to machine-translated), and it is in the language known to evaluators, which is also the ~\em{target language }\em  of the evaluated pair.

%katya: I'm not sure this is intended behavior: unless I "close" the em tag, it keeps making everything after it italic. I thought brackets should limit this...
The additional elements of the task are what we call hints, or extra sentences that help the participant to understand the main sentence. There are two types of hints: first, the \em{source}\em, which is the same sentence as \em{reference}\em, also human-produced, but in the source language of the pair. The second type is the \em{machine-translated }\em hint, which comes from the machine translation of the \em{source }\em sentence.

%katya: I don't really have a good explanation why we used source+mt tasks, except for being exhaustive. could you suggest anything?
In the course of the experiment, the following hint combinations are offered:
\begin{itemize}
\item  Reference sentence only. The participants are asked to fill the gaps not being given
any context. This task serves as a baseline score and as an indicator of phrases
that can be completed using common knowledge or language intuition (e.g.
idioms and strong collocations);
\item  Reference sentence and source sentence. By setup, the
participants have no command of the source language, however, it may help them to fill
in proper nouns or loan words;
\item  Reference sentence and machine-translated hint. In addition to the task, the
participants see the source sentence translated via Apertium. This type of task is
used for measuring the contribution of machine translation to understanding the
gist of text;
\item  Reference sentence and both the source and machine-translated hint. This task is
added for completing the possible combination of task types, and to check
for unexpected insights, should they arise.
\end{itemize}

In order to prepare the evaluation questions, we determine and remove the keywords in the reference sentences. We consider two parameters: the list of allowed parts of speech (POS), and the number of gaps relative to sentence length ("gap density"). For the evaluations described in this paper we use gap densities of 10, 20 and 30 percent, and the following parts of speech: noun, proper noun, adjective, adverb and lexical verb (as opposed to auxiliary verb).

For each sentence, the list of candidate keywords is prepared. It constitutes all the words that fall into the allowed POS list. The number of gaps in the sentence is calculated based on sentence length and specified gap density. Finally, the required number of keywords is selected from the candidate list in such a manner that the gaps are distributed evenly throughout the sentence.

Keyword removal is one of the most time-consuming steps in task preparation. In our setup, the above procedure is performed by a script integrated into the task generation pipeline. Parts of speech are determined with Apertium morphological analyzer. To control for homonymy, we only allow the word into the candidate list if all of its possible part of speech attributions are on the POS list.

Having prepared the sentence sets, we assemble them into XML formatted for the Appraise platform.

%To exclude headlines and also to allow at least one gap in 10\%-gap sentences, only the
%sentences over 10 words long are used in the tasks.
%The sentences for each experiment are drawn from various parallel text sources:
%\begin{enumerate}
%\item  For Basque-Spanish, from the corpus of legal texts Memorias de traducci\'on del
%Servicio Oficial de Traductores del IVAP 3
%\item  For English-Kazakh, from the official website of the President of the Republic of
%Kazakhstan 4
%\item  For Tatar-Russian, from the following sources on three different topics:
%  \begin{enumerate}
%    \item  Casual conversations, from a textbook 5 of spoken Tatar;
%    \item  Legal texts, from the Constitution and laws 6 of Tatarstan;
%    \item  News, from the President of Tatarstan website 7 .
%  \end{enumerate}
%\end{enumerate}
%
%Each set features 36 pairs of sentences. For the first two experiments the pairs are drawn
%randomly from the corpora, for Tatar-Russian, compiled by hand by the developer of the
%language pair. After processing, the material is transformed into XML used for uploading
%into the evaluation system (see section 2.4).

\section{Experimental set-up}
\label{sec:setup}

In this section we will discuss the evaluators, the evaluation procedure, and the tasks in more detail.

For each experiment we called for native speakers of L2 of the language pair (i.e.
Spanish, Kazakh, and Russian) who had no command of L1 of the pair (Basque, English
and Tatar, respectively). 11 evaluators participated in the in Basque-Spanish experiment, 8 in
English-Kazakh, and 28 in Tatar-Russian (although not everyone has completed their task
in full, see discussion).

%katya: maybe I should put this into the previous section?
By design, our gap-filling tasks require a human translation (reference) of source sentences. Calling for a human translator, however, would significantly increase the resources needed for evaluation. We therefore use parallel text sources, which provide the same sentence in two languages simultaneously:
\begin{enumerate}
\item  For Basque-Spanish, from the corpus of legal texts Memorias de traducci\'on del
Servicio Oficial de Traductores del IVAP\footnote{http://opendata.euskadi.net/memorias-de-traduccion-del-servicio-oficial-de-traductores-del-ivap/w79-contdata/es/};
\item  For English-Kazakh, from the official website of the President of the Republic of
Kazakhstan\footnote{http://www.akorda.kz/};
\item  For Tatar-Russian, from the following sources on three different topics:
  \begin{enumerate}
  %whoops, what is the best way to make it understand cyrillic?
    \item  Casual conversations, from a textbook\footnote{???????? ?.?. ? ??????? ???????? ??-????????. ??????: ????????? ??. ???-??, 1994. ? 320 ?. ISBN 5?298?00463?6 (???. 219, 220, 232, 233, 234)} of spoken Tatar;
    \item  Legal texts, from the Constitution and laws\footnote{http://tatarstan.ru} of Tatarstan;
    \item  News, from the President of Tatarstan website\footnote{http://president.tatarstan.ru/}.
  \end{enumerate}
\end{enumerate}

Each set features 36 pairs of sentences. For the first two experiments the pairs are drawn
randomly from the corpora, for Tatar-Russian, compiled by hand by the developer of the
language pair.

@@

The evaluations take place online, in a system called Appraise \citep{federmann12}, which is designed specifically for various MT
evaluation tasks. We adapted the code to accommodate for the gap-filling tasks. The
tasks are uploaded into the system and manually distributed between the participants by
the following rules:
\begin{enumerate}
\item  Each participant evaluates every sentence (understood as a succession of words),
a total of 36;
\item  Each participant evaluates 9 sentences in each of the four modes (see section~\ref{sec:methodology};
\item  All sentences of the set are evaluated with 10, 20 and 30\% of words removed;
\item  Each sentence-mode-percentage combination is evaluated by at least three participants.
\end{enumerate}
The participants are given the instructions in their native language above each task. The
tasks are split into smaller sets for the participants'۪ convenience. The instructions are the following: read all the available hints and fill each gap with one suitable word, guessing if unsure.
Participants'۪ answers are recorded and marked correct or incorrect automatically. In
addition, the time taken to fill the gaps in one sentence is recorded.

This variety of the gap-filling task requires open answers, and it is therefore possible that the participants may
provide words that fit the gaps well, but do not match the original answer. To account for
these cases, we process all the answers to detect possible synonyms (a method suggested by \citet{oregan13}. An answer is
considered a candidate synonym if it is given by two or more evaluators, and it does not match
the answer key. We record each candidate synonym along with the answer key and the
context sentence. Based on this data, a native speaker of the target language decides
whether the candidate synonym is indeed synonymic to the answer key in the given
context. We then check participants'۪ results against the compiled synonym list and
increase scores where appropriate. Candidate synonym extraction and score update is performed automatically.

% the sections below are from the report and have not been edited yet.

@@

The following trends may be derived from the data above:
\begin{enumerate}
\item  In general, tasks with MT assistance score higher than task without MT
assistance;
\item  The more words are removed from context, the more difficult it is to fill the gaps
(see below for exceptions);
\item  Task without hints take less time to complete than others.
\end{enumerate}
The first two trends align with our expectations. Higher scores on MT tasks indicate
better understanding of text. The large number of gaps in the sentence makes it more
difficult to predict the answer based on the context, and also leaves more room for
translation mistakes. We would expect that tasks without hints take the largest amount of
time rather than that the smallest, since the participant must come up with their own
answer unassisted. However, in the no-hint task the participant reads only one sentence,
as opposed to two or three in other tasks, which may be the source of difference.
In Basque-Spanish and Tatar-Russian these trends are defined reasonably well. In
English-Kazakh, however, the scores for source-hint task (a parallel sentence in English
for assistance) are systematically higher than for other types of tasks. In addition, a 20\%no-hint task stands out with 60\% accuracy, which is substantially higher than any other
scores for this experiment.
Having analyzed the tasks and the responses for English-Kazakh, I have found the
following possible reasons of the obtained results:
\begin{enumerate}
\item  The experiment was planned for 11 evaluators, although only 8 attempted the
tasks and 6 completed them in full. Because of this, the sentences were evaluated
once or twice, and some were not attempted at all. To reliably compare results
across different categories, all tasks must be completed;
\item  Due to an error during experiment design, some sentences are not repeated across
categories, which also suggests that we may not compare these percentages;
\item  The outlier score of 60\% is made up by three sentences. Two of these scored
unusually high. It appeared that one of them is a news headline, and the other
contains clich\'es such as "express condolences". There may also be interference
from other tasks of the set, because some of the sentences are drawn from the
same news article. We would expect the same sentences to contribute to higher
percentage in other task modes, but they were not completed there;
\item  It may be the case that the participants have at least some command of English,
which would have helped them to derive extra information in the source-hint
\end{enumerate}
It remains questionable whether we can compare results for different gap densities. In the
Basque-Spanish and Tatar-Russian experiments the 10, 20 and 30\% sets were comprised
of the same sentences. However, in each case different words were removed. It appears
that some content words are easier to fill than the others. This may explain why in
Basque-Spanish the 20\% MT tasks are completed with better accuracy than 10\% tasks.
It may be worth noting that the majority of the participants reported feeling frustrated in
the course of their work, especially while working on the no-hint tasks. 9 out of 51
participants quit the experiment before completing it.
